---
title: Selecting and Filtering
description: Selecting columns and filtering rows from a table.
engine: knitr
knitr:
  opts_chunk:
    connection: con
    engine-opts:
      target: sql.duckdb
      use_glue: true
execute:
  cache: true
sidebar_position: 2
---

:::{.callout-note}

This page is based on the chapter ["Indexing (Or Lack Thereof)"](https://kevinheavey.github.io/modern-polars/indexing.html)
of the Modern Polars book.

:::

{{< include _setup-data-1.qmd >}}

## Read the Data with PRQL

Since PRQL cannot be used to create tables, data must be read from pre-created tables in a DB.

But when PRQL is used on DuckDB, it is possible to treat Parquet and other files as tables,
so that PRQL queries can be executed without creating tables.

:::{.callout-important}

PRQL 0.8 does not support `` from `data/foo.csv` ``,
so we need to write raw SQL in [S-strings](https://prql-lang.org/book/language-features/s-strings.html)
([PRQL/prql#1535](https://github.com/PRQL/prql/issues/1535)) or
use the `read_parquet` function ([PRQL/prql#2409](https://github.com/PRQL/prql/issues/2409)).

:::

```{prql}
from (read_parquet "{{parquet_path}}")
take 5
```

## Selecting Columns and Slicing Rows

For column selection, the syntax is much the same, but the syntax for row selection using position looks different.

Note that PRQL, SQL, and R are 1-based, while Python is 0-based.

:::{.panel-tabset}

### PRQL DuckDB

```{prql}
from tab
select [Dest, Tail_Number]
take 13..16
```

### SQL DuckDB

```{glue_sql}
SELECT
  Dest,
  Tail_Number
FROM tab
LIMIT 4 OFFSET 12
```

:::{.callout-tip}

DuckDB allows SQL query starts with `FROM`,
so we can also write the query as follows:

```{.sql}
FROM tab
SELECT
  Dest,
  Tail_Number
LIMIT 4 OFFSET 12
```

:::

### dplyr R

```{r}
df |>
  select(Dest, Tail_Number) |>
  slice(13:16)
```

### Python Polars

```{python}
lf.select("Dest", "Tail_Number").head(16).tail(4).collect()
```

:::

## Filtering Rows

Since PRQL does not have something like the `IN` operator of SQL now
([PRQL/prql#993](https://github.com/PRQL/prql/issues/993)),
the `filter` statement in PRQL is a bit redundant.

:::{.panel-tabset}

### PRQL DuckDB

:::{.callout-caution}

`or` operator was changed to `||` PRQL >= 0.8.
([PRQL/prql#2170](https://github.com/PRQL/prql/issues/2170))

:::

```{prql}
from tab
filter IATA_CODE_Reporting_Airline == 'AA' || IATA_CODE_Reporting_Airline == 'DL'
take 5
```

### SQL DuckDB

```{glue_sql}
FROM tab
WHERE IATA_CODE_Reporting_Airline IN ('AA', 'DL')
LIMIT 5
```

### dplyr R

```{r}
df |>
  filter(IATA_CODE_Reporting_Airline %in% c("AA", "DL")) |>
  head(5)
```

### Python Polars

```{python}
lf.filter(pl.col("IATA_CODE_Reporting_Airline").is_in(["AA", "DL"])).head(5).collect()
```

:::

## Assigning

:::{.panel-tabset}

### PRQL DuckDB

```{prql}
let f = from_text format:json """{
  "columns": ["a", "b"],
  "data": [
    [1, 10],
    [2, 20],
    [3, 30],
    [4, 40],
    [5, 50]
  ]
}"""

from f
select [ # Note: can't use `derive` here https://github.com/PRQL/prql/issues/2392
  a,
  b = case [
    a <= 3 => b / 10,
    true => b
  ]
]
```

### SQL DuckDB

:::{.callout-caution}

DuckDB's division operator may be changed near future.
([duckdb/duckdb#7082](https://github.com/duckdb/duckdb/pull/7082))

:::

```{glue_sql}
WITH table_0 AS (
  SELECT
    1 AS a,
    10 AS b
  UNION
  ALL
  SELECT
    2 AS a,
    20 AS b
  UNION
  ALL
  SELECT
    3 AS a,
    30 AS b
  UNION
  ALL
  SELECT
    4 AS a,
    40 AS b
  UNION
  ALL
  SELECT
    5 AS a,
    50 AS b
),

f AS (
  SELECT
    a,
    b
  FROM
    table_0 AS table_1
)

SELECT
  a,
  CASE
    WHEN a <= 3 THEN b / 10
    ELSE b
  END AS b
FROM
  f
```

### dplyr R

```{r}
f <- dplyr::tibble(a = 1:5, b = a * 10)

f |>
  mutate(
    b = case_when(
      a <= 3 ~ b %/% 10,
      .default = b
    )
  )
```

### Python Polars

```{python}
f = pl.DataFrame({"a": list(range(1, 6)), "b": list(range(10, 60, 10))})

f.with_columns(
    b=pl.when(pl.col("a") <= 3).then(pl.col("b") // 10).otherwise(pl.col("b"))
)
```

:::

{{< include _cleanup-knitr.qmd >}}
